Overlay databank unlocks data-driven analyses of biomolecules for all

### 2023-10-23 main.tex
Update

\affil[1]{University of Helsinki, Institute of Biotechnology, Helsinki, Finland}
\affil[2]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\affil[3]{Department of Biomaterials, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\affil[4]{University of Potsdam, Institute of Physics and Astronomy, Potsdam-Golm, 14476, Germany}
%\affil[2]{Affiliation, department, city, postcode, country}
\affil[5]{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, Jyv{\"a}skyl{\"a}, 40014 , Finland}
\affil[6]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, Sombrerete, Zacatecas, 99102, M\'{e}xico}
\affil[7]{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg,  Halle (Saale), 06120, Germany}
\affil[8]{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), Paris, 75005, France}
\affil[9]{Universit{\'e} Paris Cit{\'e}, UFR Sciences du Vivant, Paris, 75013, France}
\affil[10]{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela,  Santiago de Compostela, E-15782, Spain}
\affil[11]{no affiliation}
\affil[12]{Institute of Biological Information Processing: Structural Biochemistry (IBI-7), Forschungszentrum Jülich, 52428 Jülich, Germany}
\affil[13]{ariadne.ai GmbH (Germany), Häusserstra{\ss}e 3 Heidelberg 69115, Germany }
\affil[14]{Department of Physical Chemistry of Drugs and Faculty of Pharmacy, Comenius University Bratislava, 832 32 Bratislava, Slovakia}
\affil[15]{Institute of Organic Chemistry and Biochemistry of the Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2,  Prague, CZ-16610, Czech Republic}
\affil[16]{School of Pharmacy, University of Eastern Finland, 70211 Kuopio, Finland}
\affil[17]{Department of Molecular Medicine, Morsani College of Medicine, University of South Florida, Tampa, Florida 33612, United States of America}
\affil[18]{Center for Global Health and Infectious Diseases Research, Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida 33612, United States of America}
\affil[19]{Department of Chemistry, University of Bergen, 5020 Bergen, Norway}
\affil[20]{Computational Biology Unit, Department of Informatics, University of Bergen, 5020 Bergen, Norway}
\affil[21]{Chemistry, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom}
\affil[22]{Department of Applied Physics, Faculty of Physics, University of Santiago de Compostela, Santiago de Compostela, E-15782, Spain}
\affil[23]{MD.USE Innovations S.L., Edificio Emprendia, 15782 Santiago de Compostela, Spain}
\affil[24]{VTT Technical Research Centre of Finland, Espoo, Finland}
\affil[25]{Hochschule Mannheim, University of Applied Sciences, Mannheim, Germany}
\affil[26]{Institut Charles Gerhardt Montpellier (UMR CNRS 5253), Universit{\'e} Montpellier, Place Eug{\`e}ne Bataillon, 34095 Montpellier, Cedex 05, France}
\affil[27]{Heidelberg University Biochemistry Center, 69120 Heidelberg, Germany}
\affil[28]{Department of Physics, University of Helsinki, FI-00014 Helsinki, Finland}
\affil[29]{IBCP CNRS, UMR 5086 7, Lyon, France}

@@

\subsubsection{Using NMRlipids databank as a training set for machine learning applications: Predicting multi-component membrane properties}

@@

\subsubsection{Extending the scope of MD simulations to new fields using NMRlipids Databank: Water diffusion anisotropy in membrane systems}
The anisotropic diffusion of water and hydrophilic molecules in directions parallel and perpendicular to membranes is important parameter in models describing the translocation of drugs through biological material, particularly in the skin \cite{hansen13,wen18,nitsche19,roberts21}. Water anisotropic diffusion plays a role also in the signal formation in diffusion tensor MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze the anisotropic diffusion of water, since only a few membrane permeation events of water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of a sufficient amount of data challenging. Here, we show that the API access to the data in NMRlipids Databank enables systematic analysis on how the anisotropic diffusion of water depends on membrane properties in multilamellar membrane systems, thereby extending the application of MD simulations to new fields.

@@

A.M.K. implemented the {\it Databank layer} and NMRlipids Databank-API.
H.S.A. contributed to the quality evaluation metrics based on NMR and X-ray scattering form factors, wrote the machine learning code, and was involved in testing the GUI, polishing the manuscript, and in community outreach to source additional data.
L.S.B. implemented roughly 630 pre-existing lipid bilayer simulations into the Databank and created the corresponding mapping files.
P.B. implemented the PCA-based equilibration metric, supervised the work of A.K., and participated in code refactoring.
F.F.-R. set up and performed simulations with Slipids Force Field: ID 617, ID 207 and ID 82. 
T.M.F. performed and analysed the solid-state NMR experiments.
P.F.J.F. contributed to changes in the buildH software so that the Databank can handle united atom simulation data.
R.G.-F. was responsible for the design, validation and supervision of the back-end and front-end for the graphical user interface (GUI) development (databank.nmrlipids.fi).
I.G. advised on implementation of the PCA-based equilibration metric and supervised the work of A.K.
B.K. contributed to the initial design and implementation of the Databank and was involved with writing the manuscript.
N.K. contributed experimental scattering data. % and SIMtoEXP algorithm.
P.K. Implemented the flip-flop analysis and supported the integration of PCA equilibration metric into the databank.
M.K. provided simulations with the OPLS3e force field. 
A.K. implemented the PCA-based equilibration metric.
A.L. provided simulations of heterogeneous lipid membranes including sphingolipids and glycosphingolipids with CHARMM36 force field.
F.S.-L. developed the back-end and participated in the design and development and validation of the graphical user interface (GUI) (databank.nmrlipids.fi).
F.L. provided simulations using the CHARMM36m force field, incorporating cholesterol POPC and SAPI25 lipids.
J.J.M. provided simulations with the Amber and CHARMM36m force fields, general discussion, and manuscript editing.
M.S.M supported O.H.S.O. in conceptualising the project, contributed to the quality evaluation metrics based on NMR and X-ray scattering form factors, assisted in refining the manuscript and the GUI, and guided the work of L.S.B. and M.K.
C.M. worked on the databank implementation and contributed code improvements.
L.M. performed simulations (for sphingomyelin, CHARMM36 force field, and Orange force field)
R.N. implemented the code for the calculations of X-ray scattering form factors.
A.N. provided simulations with TLCL lipid, simulations with Reaction-Field instead of PME, and improved the implementation.
T.J.P. performed many of the all-atom and united-atom simulations and contributed to discussions on the project.
A.P. was responsible for the design, validation and supervision of the back-end and front-end for the graphical user interface (GUI) development (databank.nmrlipids.fi).
S.S. contributed simulations of heterogeneous lipid membranes including sphingomyelin and ceramides with CHARMM36 force field.
R.T. Provided simulations of complex lipid-bilayers with charmm36m force field.
N.R. contributed simulations of complex lipid-bilayers obtained with NAMD and the charmm36 force field
O.H.S.O. designed and supervised the work, implemented the water permeation, diffusion and machine learning analyses, and wrote the manuscript.

>>>>

\affil[1]{University of Helsinki, Institute of Biotechnology, Helsinki, Finland}
\affil[2]{Department of Theory and Bio-Systems, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\affil[3]{Department of Biomaterials, Max Planck Institute of Colloids and Interfaces, 14424 Potsdam, Germany}
\affil[4]{University of Potsdam, Institute of Physics and Astronomy, Potsdam-Golm, 14476, Germany}
%\affil[2]{Affiliation, department, city, postcode, country}
\affil[5]{Nanoscience Center and Department of Chemistry, University of Jyv{\"a}skyl{\"a}, P.O. Box 35, Jyv{\"a}skyl{\"a}, 40014 , Finland}
\affil[6]{Departamento de Ciencias B\'{a}sicas, Tecnol\'{o}gico Nacional de M\'{e}xico - ITS Zacatecas Occidente, Sombrerete, Zacatecas, 99102, M\'{e}xico}
\affil[7]{NMR group - Institute for Physics, Martin Luther University Halle-Wittenberg,  Halle (Saale), 06120, Germany}
\affil[8]{Sorbonne Universit{\'e}, Ecole Normale Sup{\'e}rieure, PSL University, CNRS, Laboratoire des Biomol{\'e}cules (LBM), Paris, 75005, France}
\affil[9]{Universit{\'e} Paris Cit{\'e}, UFR Sciences du Vivant, Paris, 75013, France}
\affil[10]{Center for Research in Biological Chemistry and Molecular Materials (CiQUS), Universidade de Santiago de Compostela,  Santiago de Compostela, E-15782, Spain}
\affil[11]{no affiliation}
\affil[12]{Institute of Biological Information Processing: Structural Biochemistry (IBI-7), Forschungszentrum Jülich, 52428 Jülich, Germany}
\affil[13]{ariadne.ai GmbH (Germany), Häusserstra{\ss}e 3 Heidelberg 69115, Germany }
\affil[14]{Department of Physical Chemistry of Drugs and Faculty of Pharmacy, Comenius University Bratislava, 832 32 Bratislava, Slovakia}
\affil[15]{Institute of Organic Chemistry and Biochemistry of the Czech Academy of Sciences, Flemingovo n\'{a}m. 542/2,  Prague, CZ-16610, Czech Republic}
\affil[16]{School of Pharmacy, University of Eastern Finland, 70211 Kuopio, Finland}
\affil[17]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, 33612, United States of America}
\affil[18]{Department of Molecular Medicine, Morsani College of Medicine, University of South Florida, Tampa, Florida, 33612, United States of America}
\affil[19]{Department of Chemistry, University of Bergen, 5020 Bergen, Norway}
\affil[20]{Computational Biology Unit, Department of Informatics, University of Bergen, 5020 Bergen, Norway}
\affil[21]{Chemistry, University of Southampton, Highfield, Southampton, SO17 1BJ, United Kingdom}
\affil[22]{Department of Applied Physics, Faculty of Physics, University of Santiago de Compostela, Santiago de Compostela, E-15782, Spain}
\affil[23]{MD.USE Innovations S.L., Edificio Emprendia, 15782 Santiago de Compostela, Spain}
\affil[24]{VTT Technical Research Centre of Finland, Espoo, Finland}
\affil[25]{Hochschule Mannheim, University of Applied Sciences, Mannheim, Germany}
\affil[26]{Institut Charles Gerhardt Montpellier (UMR CNRS 5253), Universit{\'e} Montpellier, Place Eug{\`e}ne Bataillon, 34095 Montpellier, Cedex 05, France}
\affil[27]{Heidelberg University Biochemistry Center, 69120 Heidelberg, Germany}
\affil[28]{Department of Physics, University of Helsinki, FI-00014 Helsinki, Finland}
\affil[29]{IBCP CNRS, UMR 5086 7, Lyon, France}

@@

\subsubsection{Using NMRlipids Databank as a training set for machine learning applications: Predicting multi-component membrane properties}

@@

\subsubsection{Extending the scope of MD simulations to new fields using NMRlipids Databank: Water diffusion anisotropy in membrane systems}
The anisotropic diffusion of water and hydrophilic molecules in directions parallel and perpendicular to membranes is an important parameter in models describing the translocation of drugs through biological material, particularly in the skin \cite{hansen13,wen18,nitsche19,roberts21}. Water anisotropic diffusion plays a role also in the signal formation in diffusion-tensor MRI imaging~\cite{topgaard20}. MD simulations are rarely used to analyze the anisotropic diffusion of water, since only a few membrane permeation events of water are typically observed in a single MD simulation trajectory~\cite{venable19,camilo2022}, thereby making the collection of a sufficient amount of data challenging. Here, we show that the API access to the data in NMRlipids Databank enables systematic analysis on how the anisotropic diffusion of water depends on membrane properties in multilamellar membrane systems, thereby extending the application of MD simulations to new fields.

@@

A.M.K. implemented the {\it Databank layer} and NMRlipids Databank-API.
H.S.A. contributed to the quality evaluation metrics based on NMR and X-ray scattering form factors, wrote the machine learning code, and was involved in testing the GUI, polishing the manuscript, and in community outreach to source additional data.
L.S.B. implemented roughly 630 pre-existing lipid bilayer simulations into the Databank and created the corresponding mapping files.
P.B. implemented the PCA-based equilibration metric, supervised the work of A.K., and participated in code refactoring.
F.F.-R. set up and performed simulations with Slipids Force Field: ID 617, ID 207 and ID 82. 
T.M.F. performed and analysed the solid-state NMR experiments.
P.F.J.F. contributed to changes in the buildH software so that the Databank can handle united atom simulation data.
R.G.-F. was responsible for the design, validation and supervision of the back-end and front-end for the graphical user interface (GUI) development (databank.nmrlipids.fi).
I.G. advised on implementation of the PCA-based equilibration metric and supervised the work of A.K.
B.K. contributed to the initial design and implementation of the Databank and was involved with writing the manuscript.
N.K. contributed experimental scattering data. % and SIMtoEXP algorithm.
P.K. Implemented the flip-flop analysis and supported the integration of PCA equilibration metric into the databank.
M.K. provided simulations with the OPLS3e force field. 
A.K. implemented the PCA-based equilibration metric.
A.L. provided simulations of heterogeneous lipid membranes including sphingolipids and glycosphingolipids with CHARMM36 force field.
F.S.-L. developed the back-end and participated in the design and development and validation of the graphical user interface (GUI) (databank.nmrlipids.fi).
F.L. provided simulations using the CHARMM36m forcefield, incorporating cholesterol POPC and SAPI25 lipids.
J.J.M. provided simulations with the Amber force field, general discussion, and manuscript editing.
M.S.M supported O.H.S.O. in conceptualising the project, contributed to the quality evaluation metrics based on NMR and X-ray scattering form factors, assisted in refining the manuscript and the GUI, and guided the work of L.S.B. and M.K.
C.M. worked on the databank implementation and contributed code improvements.
L.M. performed simulations (for sphingomyelin, CHARMM36 force field, and Orange force field)
R.N. implemented the code for the calculations of X-ray scattering form factors.
A.N. provided simulations with TLCL lipid, simulations with Reaction-Field instead of PME, and improved the implementation.
T.J.P. performed many of the all-atom and united-atom simulations and contributed to discussions on the project.
A.P. was responsible for the design, validation and supervision of the back-end and front-end for the graphical user interface (GUI) development (databank.nmrlipids.fi).
S.S. contributed simulations of heterogeneous lipid membranes including sphingomyelin and ceramides with CHARMM36 force field.
R.T. Provided simulations of complex lipid-bilayers with charmm36m force field.
N.R. contributed simulations of complex lipid-bilayers obtained with NAMD and the charmm36 force field
O.H.S.O. designed and supervised the work, implemented the water permeation, diffusion and machine learning analyses, and wrote the manuscript.
